White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-12-17 papers/wdmerger_I/paper.tex
Rearrange figure placement

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that initializes a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used \castro\ to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6\ \text{cm}, 1.6\ \text{cm}]^3$, and
we take $R = 1.0\ \text{cm}$ and $\rho = 10^3\ \text{g cm}^{-3}$. 
The zones with $r > R$ are filled with an ambient material of very low density 
($10^{-8}\ \text{g cm}^{-3}$). We run this problem at multiple 
resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. We measure the 
numerical error by calculating the $L^2$ norm of the error and 
normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}
We define the order of convergence $p$ between two simulations with a jump 
in resolution of integer factor $m > 1$ as
\begin{equation}
  p = \text{log}_{m}\left(\frac{\text{Error}_{\text{low}}}{\text{Error}_{\text{high}}}\right).
\end{equation}
Here $\text{Error}_{\text{low}}$ is the $L^2$ error at the lower resolution 
and $\text{Error}_{\text{high}}$ is the $L^2$ error at the higher resolution.
We expect the error to converge at $p = 2$ given the discretization we choose. 
For all simulations in this section and for all our main science simulations,
we choose a relative error tolerance of $10^{-10}$ to be satisfied in the multigrid solve.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/phi_comparison}
  \caption{Comparison of the \castro\ gravitational potential to the analytical solution for: 
    a sphere of uniform density; the same sphere, but with the potential normalized using the 
    actual amount of mass on the grid instead of the mass of a perfect sphere; and, a 
    cube of uniform density. Plotted also is a notional curve whose slope represents
    perfect second order convergence.\label{fig:gravity_convergence}}
\end{figure}
The results of this test are plotted in \autoref{fig:gravity_convergence}.

We find that at low resolution convergence is actually substantially better 
than second-order. The explanation for this is that we are attempting to 
model a spherical object on a rectangular grid. This results in two sources of error.
First, at very low resolution, the object does not look very spherical due to the rectangular 
grid representation, so the potential it produces is not quite that of a sphere. 
As the resolution is increased, the distribution of the mass on the grid will change.
Second, the total amount of mass on the grid will change as the sphere fills out. 
So we are combining the true accuracy bonus from increased resolution 
with the artificial accuracy bonus from getting closer to solving the problem 
we are supposed to be solving. At high resolution this effect levels off, though, 
as the representation of the sphere is not significantly different in 
our two highest resolutions shown. For example, at $128$ zones per dimension 
the amount of mass on the grid happens to be slightly closer to the true spherical 
mass than at $256$ zones per dimension.

We can eliminate the second source of error by changing the density on  
the grid so that the total mass $M$ is actually what we intend it to be.
The resolution study for this case (the ``normalized sphere'') is also
plotted in \autoref{fig:gravity_convergence}. At low resolution we still obtain
convergence slightly better than second-order, indicating that we 
have not eliminated the geometrical problem of the mass distribution changing.

>>>>

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that initializes a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used \castro\ to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6\ \text{cm}, 1.6\ \text{cm}]^3$, and
we take $R = 1.0\ \text{cm}$ and $\rho = 10^3\ \text{g cm}^{-3}$. 
The zones with $r > R$ are filled with an ambient material of very low density 
($10^{-8}\ \text{g cm}^{-3}$). We run this problem at multiple 
resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. We measure the 
numerical error by calculating the $L^2$ norm of the error and 
normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}
We define the order of convergence $p$ between two simulations with a jump 
in resolution of integer factor $m > 1$ as
\begin{equation}
  p = \text{log}_{m}\left(\frac{\text{Error}_{\text{low}}}{\text{Error}_{\text{high}}}\right).
\end{equation}
Here $\text{Error}_{\text{low}}$ is the $L^2$ error at the lower resolution 
and $\text{Error}_{\text{high}}$ is the $L^2$ error at the higher resolution.
We expect the error to converge at $p = 2$ given the discretization we choose. 
For all simulations in this section and for all our main science simulations,
we choose a relative error tolerance of $10^{-10}$ to be satisfied in the multigrid solve.
The results of this test are plotted in \autoref{fig:gravity_convergence}.

We find that at low resolution convergence is actually substantially better 
than second-order. The explanation for this is that we are attempting to 
model a spherical object on a rectangular grid. This results in two sources of error.
First, at very low resolution, the object does not look very spherical due to the rectangular 
grid representation, so the potential it produces is not quite that of a sphere. 
As the resolution is increased, the distribution of the mass on the grid will change.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.45]{plots/phi_comparison}
  \caption{Comparison of the \castro\ gravitational potential to the analytical solution for: 
    a sphere of uniform density; the same sphere, but with the potential normalized using the 
    actual amount of mass on the grid instead of the mass of a perfect sphere; and, a 
    cube of uniform density. Plotted also is a notional curve whose slope represents
    perfect second order convergence.\label{fig:gravity_convergence}}
\end{figure}
Second, the total amount of mass on the grid will change as the sphere fills out. 
So we are combining the true accuracy bonus from increased resolution 
with the artificial accuracy bonus from getting closer to solving the problem 
we are supposed to be solving. At high resolution this effect levels off, though, 
as the representation of the sphere is not significantly different in 
our two highest resolutions shown. For example, at $128$ zones per dimension 
the amount of mass on the grid happens to be slightly closer to the true spherical 
mass than at $256$ zones per dimension.
We can eliminate the second source of error by changing the density on  
the grid so that the total mass $M$ is actually what we intend it to be.
The resolution study for this case (the ``normalized sphere'') is also
plotted in \autoref{fig:gravity_convergence}. At low resolution we still obtain
convergence slightly better than second-order, indicating that we 
have not eliminated the geometrical problem of the mass distribution changing.
